---
title: "2. R Programming: Week 3 Assignment"
output: html_notebook
---

# Intro
This is the matrix inversion cache assignment. The repo containing the starter scripts is found [here](https://github.com/rdpeng/ProgrammingAssignment2). I'm not cloning that one, I'm just gonna do my own thing here.

I've gotta write 2 functions, `makeCacheMatrix()` and `cacheSolve()`.

* `makeCacheMatrix`: creates a special matrix object that can cache its own inverse
* `cacheSolve`: computes the inverse of the special object created by `makeCacheMatrix` by checking for a cached value and then calculating the inverse only if necessary

Once the functions are written, I need to write some test cases to prove that it works.

## `makeCacheMatrix`
This function needs to:

* set the value of the matrix
* get the value of the matrix
* set the value of the inverse
* get the value of the inverse

... and then store those functions in a list.

```{r}
makeCacheMatrix <- function(x = numeric()) {
     
     cinv <- NULL
     
     set <- function(y) {
          x <<- y
          cinv <<- NULL
     }
     
     get <- function() x
     
     setinverse <- function(inverse) cinv <<- inverse
     
     getinverse <- function() cinv
     
     list(set = set, get = get, setinverse = setinverse, getinverse = getinverse)
}
```

## `cacheSolve`
This one needs to:

* get the cached inverse value
* if that's null, calculate the inverse
* then set the inverse

```{r}
cacheSolve <- function(x, ...) {
     inv <- x$getinverse()
     if(!is.null(inv)) {
          message("getting cached data")
          return(inv)
     }
     data <- x$get()
     inv <- solve(data, ...)
     x$setinverse(inv)
     inv
}
```

## Tests
This section just runs some tests to demonstrate that this thing is actually caching and retrieving results. First, we make a matrix and use the `makeCacheMatrix` function on it. Then we can use `cacheSolve` to find the inverse.

```{r}
set.seed(75)
mtx <- matrix(rnorm(25, 100, 10), 5, 5)
newmtx <- makeCacheMatrix(mtx)
solve(mtx)
cacheSolve(newmtx)
```

So we managed to solve correctly for the inverse. Kewl.

Now, if we ask for the inverse again, we should just pull from the cache, rather than recomputing it. Let's see.

```{r}
cacheSolve(newmtx)
```

Damn it feels good to be a gangsta.
