---
title: "2. R Programming: Week 4 Assignment"
output: html_notebook
---

# Intro
This is the hospital assignment. The PDF and dataset are in the data directory of this week. The goal is to sort/arrange/subset/filter a dataset. Let's begin.

## 1. plot the data
First, we'll read in the outcome data, then we'll take a look at a small portion of it.

```{r}
dir("./data")
outcome <- read.csv(paste0("./data/", dir("./data")[3]), colClasses = "character")
head(outcome)
str(outcome)
```

Let's start with a histogram of the 30 day death rates from heart attacks. The "outcomes" dataset lists > 4000 hospitals, and column 11 is the 30-day risk-adjusted rate for heart attacks. We'll coerce this column to be numeric and make a histogram.

```{r}
outcome[, 11] <- as.numeric(outcome[, 11])
hist(outcome[, 11])
```

## 2. find the best hospital in the state
Now we'll write a function `best()` that will take a state and outcome name and return the top hospital. Need to return the `Hospital.Name`. Outcome can be either "heart attack", "heart failure", or "pneumonia." Exclude `NA`'s. Return 1st hospital alphabetically in the event of a tie. Here's the plan:

1. read the .csv
2. pull only relevant columns
3. make list of states (from data)
4. make list of outcomes
5. rename 3 outcomes columns
6. check for valid state and outcome input
7. split the dataset by state
8. order each state subset by hospital name (for ties)
9. use `which.min()` to find the index of the minimum value for this outcome
10. use that index to return the correct hospital name

```{r}
best <- function(state, outcome) {
     
     # read the outcome data, cast outcomes as numeric
     all <- read.csv(paste0("./data/", dir("./data")[3]), colClasses = "character")
     data <- all[, c(2, 7, 11, 17, 23)]
     
     # generate vector of valid states based on state abbrreviations from the data
     # make vector of outcomes
     states <- levels(as.factor(data[, 2]))
     outcomes <- c("heart attack", "heart failure", "pneumonia")
     
     for (c in 3:5) {
          suppressWarnings(data[, c] <- as.numeric(data[, c]))
          names(data)[c] <- outcomes[c-2]
     }
     
     # check that the state and outcome inputs are valid
     if (!is.element(outcome, outcomes)) {
          stop("invalid outcome")
     } else if (!is.element(state, states)) {
          stop("invalid state")
     }
     
     # split data by state; order every data frame in list by hospital name
     data <- lapply(split(data, data$State), function(x) x[order(x$Hospital.Name), ])
     
     # find index of outcome minimum for each state, then for this state
     indvec <- sapply(data, function(x) which.min(x[[outcome]]))
     ind <- unname(indvec[state])
     
     # return the hospital name in that state whith the lowest 30-day death rate
     name <- as.vector(data[[state]]$Hospital.Name)[ind]
     
     name
}
```

Now, we test:

```{r, error=TRUE}
best("TX", "heart attack")
best("TX", "heart failure")
best("MD", "heart attack")
best("MD", "pneumonia")
best("BB", "heart attack")
best("NY", "hert attack")
```

## 3. Rank hospitals by outcome in a state
I need to write a function called `rankhospital` that takes 3 args: state, outcome, the ranking of a hospital in that state for that outcome. I need to return the hospital name that has that rank. The rank arg can accept "best" or "worst" as well as integers; return `NA` if the number > # of hospitals in that state. Exclude `NA`'s from rankings. Ties should be broken by hospital name.

Need to check the validity of args.

The plan:

1. if rank is "best", call `best()`
2. read.csv
3. pull only relevant columns
4. make list of states (from data)
5. make list of outcomes
6. rename 3 outcome columns
7. check for valid state and outcome input
8. if rank is "worst"
     - split data by state, order by outcome (desc) and hospital name (asc)
     - use `which.max` to find the index of the max value for this outcome (spoiler, it's the 1st in the list, cause we ordered by this)
     - use that index (1) to return hospital name
9. else
     - split data by state, order by outcome (asc)
     - use `num` arg as index to return hospital name
     
     
```{r}
rankhospital <- function(state, outcome, num = "best") {
     
     if (num == "best") {
          name <- best(state, outcome)
     }
     
     data <- read.csv(paste0("./data/", dir("./data")[3]), colClasses = "character")
     data <- data[, c(2, 7, 11, 17, 23)]
     
     states <- levels(as.factor(data[, 2]))
     outcomes <- c("heart attack", "heart failure", "pneumonia")
     
     for (c in 3:5) {
          suppressWarnings(data[, c] <- as.numeric(data[, c]))
          names(data)[c] <- outcomes[c-2]
     }
     
     if (!is.element(outcome, outcomes)) {
          stop("invalid outcome")
     } else if (!is.element(state, states)) {
          stop("invalid state")
     }
     
     if (num == "worst") {
          wdata <- lapply(split(data, data$State), function(x) x[order(-x[outcome], x$Hospital.Name), ])
          indvec <- sapply(wdata, function(x) which.max(x[[outcome]]))
          ind <- unname(indvec[state])
          name <- as.vector(wdata[[state]]$Hospital.Name)[ind]
     } else {
          rdata <- lapply(split(data, data$State), function(x) x[order(x[outcome], x$Hospital.Name), ])
          ind <- num
          name <- as.vector(rdata[[state]]$Hospital.Name)[ind]
     }
     name
}
```

And now we test.

```{r, error=TRUE}
rankhospital("TX", "heart failure", 4)
rankhospital("MD", "heart attack", "worst")
rankhospital("MN", "heart attack", 5000)
```

Nailed it.

## 4. Rank hospitals in all state
`rankall()` takes 2 args: outcome and rank (num). This function returns a 2 column data frame (state and hospital name). Column 1 is hospital name, column 2 si state. Handle ties alphabetically.

```{r}
rankall <- function(outcome, num = "best") {
     outcomes <- c("heart attack", "heart failure", "pneumonia")
     
     if (!is.element(outcome, outcomes)) {
          stop("invalid outcome")
     }
     
     data <- read.csv(paste0("./data/", dir("./data")[3]), colClasses = "character")
     data <- data[, c(2, 7, 11, 17, 23)]
     
     states <- levels(as.factor(data[, 2]))
     
     for (c in 3:5) {
          suppressWarnings(data[, c] <- as.numeric(data[, c]))
          names(data)[c] <- outcomes[c-2]
     }
     
     sp <- split(data, data$State)
     
     if (num == "worst") {
          num <- 1L
          wdata <- lapply(sp, function(x) x[order(-x[outcome], x$Hospital.Name), ])
          rdf <- data.frame(hospital = unname(sapply(wdata, function(x) x[[1]][num])), state = names(wdata))
     } else {
          rdata <- lapply(sp, function(x) x[order(x[outcome], x$Hospital.Name), ])
          
          if (num == "best") {
               num <- 1L
          }
          
          rdf <- data.frame(hospital = unname(sapply(rdata, function(x) x[[1]][num])), state = names(rdata))
     }
     rdf
}
```

And the tests.

```{r}
head(rankall("heart attack", 20), 10)
tail(rankall("pneumonia", "worst"), 3)
tail(rankall("heart failure"), 10)
```

Yep, I'm awesome.
