---
title: "3. Get and Clean Data: Week 2 Notes            "
output: html_notebook
---

# Data Sources

## Reading from MySQL
Databases, tables, fields, records. MySQL is free and commonly used in interweb applications.

So a database consists of lots of tables. A table would be a data frame in R, and the database tables each describe one logical component of the data. They're linked together by keys (id's).

To really test this, it's best to start a local installation of MySQL and run against that. For these examples we'll be using a genomic database provided by UC Santa Cruz.

You need to install RMySQL. Once you do that, you can connect to the database with:
```{r}
library(RMySQL)
hg19 <- dbConnect(MySQL(), user="genome", db="hg19", host="genome-mysql.cse.ucsc.edu")
allTables <- dbListTables(hg19)
length(allTables)

## now we can see the fields in a specific table and count the rows:
dbListFields(hg19, "affyU133Plus2")
dbGetQuery(hg19, "select count(*) from affyU133Plus2")

## you can extract a table into a data frame
affyData <- dbReadTable(hg19, "affyU133Plus2")
```

There's a lot of nuance in the RMySQL package, don't have time to unpack everything here.

## Reading from HDF5
HDF5 is used for storing large amounts of data in a hierarchical format.

- **groups** contain zero or more datasets and metadata
     - have a group header with group name and list of attributes
     - have a group symbol table with a list of objects in the group
- **datasets** are multidimensional arrays of data elements with metadata
     - have a header with a name, data type, dataspace, and storage layout
     - have a data array with the data

The R HDF5 libraries are installed through bioconductor, but they're conveniently included in this dockerfile (hopefully). Uh-firmative, BioConductor capabilities verified.

```{r}
library(rhdf5)
created = h5createFile("example.h5")
created
```

There's an RHDF5 tutorial on bioconductor.org that is the basis for this lecture.

Once we've created the `.h5` file, we can create groups within that file.

```{r}
created = h5createGroup("example.h5", "foo")
created = h5createGroup("example.h5", "baa")
created = h5createGroup("example.h5", "foo/foobaa")
h5ls("example.h5")
```

There you can see the groups created, structed like the linux filesystem. Then you can write to those groups.

```{r}
A = matrix(1:10, nrow = 5, ncol = 2)
h5write(A, "example.h5", "foo/A")
B = matrix(seq(0.1, 2.0, by = 0.1))
dim(B) <- c(5, 2, 2)
attr(B, "scale") <- "liter"
h5write(B, "example.h5", "foo/foobaa/B")
h5ls("example.h5")
```

You can also just write a data frame to a `.h5` (ps, this is what you're pretty much always do...).

```{r}
df = data.frame(1L:5L, seq(0, 1, length.out = 5), c("ab", "cde", "fghi", "a", "s"), stringsAsFactors = FALSE)
h5write(df, "example.h5", "df")
h5ls("example.h5")
```

We just added the `data.frame` object `df` to at the `/` location.

And you read these things with, surprise, `h5read`.

```{r, eval=FALSE}
readA = h5read("example.h5", "foo/A")
readB = h5read("example.h5", "foo/foobaa/B")
readdf= h5read("example.h5", "df")
```

h5 is also good for reading in chunks. We can modify our 5x2 matrix in `example.h5` with `h5write(c(12, 13, 14), "example.h5", "foo/A", index=list(1:3), 1))`

## Reading from the web
WEBSCRAPING!!! Programmatically extracting data from the tubes via the HTML that makes the tubes.

First, read in the HTML, then we parse.

```{r}
con = url("http://scholar.google.com/citations?user=HI-I60AAAAJ&hl=en")
htmlcode = readLines(con)
close(con)
htmlcode

library(xml2)
url <- "http://scholar.google.com/citations?user=HI-I60AAAAJ&hl=en"
html <- xml2::read_html(url)
html
```

