---
title: '2. R Programming: Week 1 Notes'
output:
  html_notebook: default
  html_document: default
---

# R Programming - Objects & Basics

## R Objects
R has 5 basic, atomic classes of objects:

* character
* numeric
* integer
* complex
* logical

The most basic object is a vector. A vector can only contain 1 class. For multi-class, you use a list, which is represented like a vector, but can hold multiple classes as elements.

### Numbers
Numbers are pretty much always numeric class. (But you can force integer numbers with the `L` suffix, e.g. `1L`.)

`Inf` is a special number representing infinity, and `NaN` represents undefined (not a number).

### Vectors and lists
`c()` concatenates stuff into vectors. `:` is a shortcut for sequences. `vector()` makes empty vectors.

Examples:

```{r}
x <- c(.5, .5); x
x <- c(TRUE, FALSE); x
x <- c(T, F); x
x <- c("a", "b", "c"); x
x <- c(1+0i, 2+4i); x
x <- 9:17; x
x <- vector("numeric", length = 10); x
```

If you mix classes, R will coerce a vector into the 'least common denominator' class, a class that can represent all elements. Examples:

```{r}
x <- c(1.7, "a"); class(x)
x <- c(TRUE, 2); class(x)
x <- c("a", TRUE); class(x)
```

You can force changes with `as.numeric()`, `as.logical()`, `as.character()`. Non-sensible coercion will result in `NA` values.

Lists are special types of vectors that can hold elements of different classes. Lists can also hold vectors, so the first element of a list can be a vector with length > 1.

```{r}
x <- list(1, "a", TRUE, 1+4i)
lapply(x, class)
```

### Matrices
Matrices are like vectors, but they have a dimension attribute. The dimension attribute, itself, is just an integer vector of length 2, `(nrows, ncols)`.

`matrix(f, r, c)` makes a matrix with `f` in each element--filled from the top left, then down, then right--and `r` rows by `c` columns.

You can also turn a vector into a matrix by passing the dimensions to the `dim` attribute of the vector.

Finally, you can `cbind()` and `rbind()` together multiple vectors (of the same length). This just tacks them together as columns/rows.

Examples:
```{r}
m <- matrix(1:12, 4, 3); m

m <- 1:21
dim(m) <- c(3, 7); m

x <- 1:6
y <- 10:16
cbind(x, y)
rbind(x, y)
```

### Factors
Factors represent categorical data. It's basically an integer vector where each integer corresponds to a label (like male/female). They can be ordinal or just categorical. Factors are treated specially by modeling functions like `lm()` or `glm()`.

Factors are created with the `factor()` function, input is a character vector. The factor contains a level for each category of data; those levels are actually stored in R as integers. Levels are applied alphabetically, so in the below example, `no` is assigned 1 and `yes` is assigned to 2.

```{r}
x <- factor(c("yes", "no", "no", "yes", "no", "yes")); x
```

You can force the order of levels like so:
```{r}
x <- factor(c("yes", "no", "no", "yes", "no", "yes"), levels = c("yes", "no")); x
```

### Missing values
Missing values are either `NA` or `NaN`. `NA` is any missing value, `NaN` is for undefined mathematical operations. `NA`'s can have a class, so there can be integer/character/numeric variations. `NaN` values are technically also `NA`, but the converse is not true.

### Data frames
Data frames are a special type of list where every element of the list must be of the same length; they're used to represent tabular data. Each element of the list can have a different class, unlike matrices.

Data frames have a special attribute called `row.names` which identifies each row. They can be created by calling `read.table()` or `read.csv()`, and they can be converted to a matrix with `data.matrix()`. This conversion will force a least common denominator approach regarding object class.

### The names attribute
R objects can have names, assigned via the `names()` function. Names can also be applied directly in the creation of an object.

```{r}
x <- 1:3
names(x) <- c("foo", "bar", "norf"); x
x <- list(a = 1, b = "g", c = TRUE); x
```

Matrices can have names as well, but these are known as `dimnames`. 

```{r}
m <- matrix(1:9, 3, 3); m
dimnames(m) <- list(c("w1", "w2", "w3"), c("h1", "h2", "h3")); m
```

## Reading tabular data

* `read.table()` and `read.csv()` read in tabular data and return a data frame (inverse of `write.table()`)
* `read.lines()` reads lines of a text file and returns a character vector (inverse of `write.lines()`)
* `source()` reads in R code (inverse of `dump()`)
* `dget()` reads in R code for R objects that have been deparsed into text files (inverse of `dput()`)
* `load()` reads in saved workspaces (inverse of `save()`)
* `unserialize()` reads in single R objects in binary form (inverse of `serialize()`)

### `read.table`
This is the most common method of reading data into R. Here are some important arguments:

* `file` is the name of the file or connection
* `header` is a logical indicating if the file has a header line
* `sep` is a string indicating how the columns are separated
* `colClasses` is a character vector indicating the class of each column
* `nrows` is the number of rows to read in
* `comment.char` is a character string indicating the comment character
* `skip` is the number of lines from the beginning to skip
* `stringsAsFactors` is a logical indicating that character variables should be coded as factors

For small/moderate datasets, you probably don't need any arguments (assuming the data is pretty clean). R automatically skips `#` lines as comments, calculates the number of rows and memory allocation, and determines the column classes (although explicitly declaring these is a good idea, speeds up the process).

### Reading larger datasets
Definitely refer to the `read.table` help page, it provides lots of good info about optimizing the function.

You should make a rough calculation of the memory requirement to make sure your machine can pull it off. For each column do $length \times bytes \space for \space data \space type$. Then divide those bytes by $2^{20} bytes/MB$ resulting in total size in MB. Add up all columns and that's the memory requirement. You'll need about twice that for things to go smoothly.

If there are no comments, then set `comment.char = ""`, speeds things up.

`colClasses` is super important. R tries to manually determine the class of each column, but that can be expensive. Good MO is to read in 100 rows, let R determine column classes, then pipe that character vector into the reading of the whole dataset.

```{r, eval=FALSE}
tenrows <- read.table(".txt", nrows = 10)
classes <- sapply(tenrows, class)
data <- read.table(".txt", colClasses = classes)
```

## Textual data formats
These are the formats for storing data that include certain metadata about the R objects they describe. They are text files, which adheres to the Unix philosophy, they work well with git, and are edit-able and potentially recoverable. However, they're not particularly space-efficient.

`dput()` allows you to deparse an R object and then read it back in via `dget()`. This process essentially writes R code that will reconstruct the R object.

```{r}
y <- data.frame(a=1, b="a")
dput(y)
dput(y, file="y.R")
new.y <- dget("y.R"); new.y
```

`dump()` is like `dput()`, but it can handle deparsing multiple R objects, which can then be read back in with `source()`.

```{r}
x <- "foo"
y <- data.frame(a=1, b="a")
dump(c("x", "y"), file = "stuff.R")
rm(x, y)
source("stuff.R")
x
y
```

## Connections
How R interfaces with the outside world.

* `file()` opens a connection to a file
     * `str(file)` shows the structure. `description` is the name of the file, `open` is a code (`r` for read only, `w` for write and initialize a new file, `a` for appending, and `rb`, `wb`, `ab` for binary mode (Windows))
* `gzfile()` opens a connection to a gzip compressed file
* `bzfile()` opens a connection to a bzip2 compressed file
* `url()` opens a connection to a webpage

In general, you don't have to deal with connections much, but FYI, you use connections according to this process:

```{r, eval=FALSE}
conn <- gzfile("file.gz")
x <- readLines(conn, 10)
close(conn)
```

## Subsetting

* `[]` always returns an object of the same class as the original object; it can also be used to extract more than one element of an object (with 1 exception)
* `[[]]` is used to extract elements of lists/data frames; it can only extract a single element, but that element can be of a different class than the original object
* `$` is used to extract elements of a list/data frame by name

#### Basic [] indexing

```{r}
x <- c("a", "b", "c", "c", "d", "a")
x[1]
x[2]
x[1:4]
```

#### Logical indexing

```{r}
x[x > "a"]
u <- x > "a"
u
x[u]
```

#### Subsetting lists

```{r}
x <- list(foo = 1:4, bar = 0.6)
x[1]           # returns a list with 1 element named foo containing 1:4
x[[1]]         # returns a vector with the sequence 1:4
x$bar          # returns the element named bar
x[["bar"]]     # same as x$bar
x["bar"]       # returns a list (forced to, by []), with the element 'bar' in it
```

Using `[]` to index multiple elements

```{r}
x <- list(foo = 1:4, bar = 0.6, baz = "hello")
x[c(1, 3)]     # returns a list with the elements foo and baz
```

Using `[[]]` to index with computed names

```{r}
name <- "foo"  # you can store an index in a variable
x[[name]]      # returns the sequence 1:4, processes name = foo
x$name         # returns NULL, does NOT evaluate name = foo
x$foo          # returns 1:4, as expected
```

Using `[[]]` to recursively index into lists

```{r}
x <- list(a = list(10, 12, 14), b = c(3.14, 2.81))
x[[c(1, 3)]]   # indexes the 3rd element of the 1st element, 14 in this case
x[[1]][[3]]    # same as above, 3rd element of 1st element, 14
x[[c(2, 1)]]   # returns 3.14, 1st element of 2nd element
```

#### Subsetting matrices
Basically works as you'd expect: rows then columns, just like the $i, j$ notation.

```{r}
x <- matrix(1:6, 2, 3)
x[1, 2]        # 1st row, 2nd col
x[2, 1]        # 2nd row, 1st col
x[1, ]         # entire 1st row
x[, 2]         # entire 2nd col
```

Note: despite the rules of the `[]` operator, when retrieving a single element of a matrix, or a single row or column, you *DO NOT* get a matrix back; a single element is returned as a vector of length 1, rather than a 1x1 matrix. You can call `x[1, 2, drop = FALSE]` to force a 1x1 matrix. Similarly, rows and columns are returned as vectors.

#### Partial matching
Partial matching of names works with the `[[]]` and `$` operators.

```{r}
x <- list(aardvark = 1:5)
x$a            # returns aardvark
x[["a"]]       # returns null unless you specify exact = FALSE
```

#### Removing NA values

```{r}
x <- c(1, 2, NA, 4, NA, 5)
bad <- is.na(x); bad
x[!bad]
```

The `complete.cases()` function generates a logical vector where the elements of all inputs are present. This also works on data frames, returning rows with complete data.

## Vectorized operations
Instead of looping, use vectorization! `x + y` will add those 2 vectors element-wise. This is true for most arithmetic operations in R.

For matrices, note that the arithmetic operations are still element-wise, and not matrix operations. True matrix multiplication requires the `%*%` operator.
