---
title: "4. Exploratory Analysis - Week 1 Assignment"
output: html_notebook
---

# The Data
This data comes from the UC Irvine Machine Learning Repo; it's available [here](https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2Fhousehold_power_consumption.zip). [Here's](https://github.com/rdpeng/ExData_Plotting1) the github page for this project.

We're looking at residential power consumption rates. 1 minute sampling interval over nearly 4 years. 2M rows and 9 columns.

## Gitkraken 'Out of Memory'
Gitkraken is throwing an error when I try to stage the data .txt file for this assignment. So for now, the downloading and placement of the file will have to take place external to git. Download the `household_power_consumption.txt` file from above, and place it in the `4. Explore/wk1/data` directory. Then source `4. Explore/wk1/project.R` and it'll complete the assignment.

# Install/load
First, we do the typical install game...

```{r installpkgs, echo=FALSE, eval=FALSE}
installedList <- rownames(installed.packages())

## It's important to note the order of pkgList. Packages should be specified in the order that they should be loaded
pkgList <- c("dplyr")
loadlist <- pkgList

toInstall <- pkgList[!is.element(pkgList, installedList)]

n <- length(toInstall)

if(n > 0) {
     install.packages(toInstall)
     print("The following packages were installed:")
     toInstall
} else {
     print("nothing to install; job's finished.")
}
```

Then we do the typical package load game...

```{r load}
library(dplyr)
```

```{r loadpkgs, echo=FALSE, eval=FALSE}
loadlist <- c("dplyr")
knitrlist <- c("ezknitr", "knitr")
otherpkgslist <- names(sessionInfo()$otherPkgs)
detachlist <- otherpkgslist[!is.element(otherpkgslist, knitrlist)]
prefixdetach <- sapply(detachlist, function(x) paste0("package:", x), USE.NAMES = FALSE)

invisible(sapply(prefixdetach, function(x) try(detach(x, character.only = TRUE))))

suppressMessages(sapply(loadlist, library, character.only = TRUE))
```

# Download
Then we do the typical download game...

```{r download, echo = FALSE}
## setup URL and path variables for this data
url <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2Fhousehold_power_consumption.zip"
datapath <- "./data/"
zippath <- paste0(datapath, "temp.zip")

## create the data directory if necessary
if(!dir.exists(datapath)) {
     dir.create(datapath)
     print("Created /data directory")
} else {
     print("/data directory already exists")
}

## If there's no data, and no .zip, then download and unzip data
## If there's no data, but there is a .zip, then unzip data
if(!length(dir(datapath)) > 0) {
     if(!file.exists(zippath)) {
          download.file(url, destfile = zippath)
          unzip(zippath, exdir = datapath)
          file.remove(zippath)
          print("Downloaded & extracted data file")
     } else {
          unzip(zippath, exdir = datapath)
          file.remove(zippath)
          print("Found .zip, extracted")
     }
} else {
     print("Found data, ready to go")
}
```

# The actual assignment
Now, we do the actual work.

```{r}
# read .txt data file
alldata <- read.table(
      "./data/household_power_consumption.txt",
      header = TRUE,
      sep = ";",
      na.strings = "?",
      colClasses = c(rep("character", 2),rep("numeric", 7))
)

figpath <- "./output-assignment/figures/"

# prepare dataset w/ dplyr
alldata <- alldata %>%
      mutate(datetime = as.POSIXct(paste(Date, Time),
          format = "%d/%m/%Y %H:%M:%S")) %>%
      select(datetime, 3:9) %>%
      filter(datetime >= "2007-02-01" & datetime < "2007-02-03")
```

```{r plot1}
# make plot 1
# png(paste0(figpath, "plot1.png"))
hist(alldata$Global_active_power,
     main = "Global Active Power",
     col = "Red",
     xlab = "Global Active Power (kilowatts)")
# dev.copy(png, file = paste0(figpath, "plot1.png"))
# dev.off()
```

```{r plot2}
# make plot 2
# png(paste0(figpath, "plot2.png"))
plot(alldata$datetime, alldata$Global_active_power,
     ylab = "Global Active Power (kilowatts)", xlab = "", type = "l")
# dev.off()
```

```{r plot3}
# make plot 3
# png(paste0(figpath, "plot3.png"))
plot(alldata$datetime, alldata$Sub_metering_1, type = "l",
     ylab = "Energy sub metering", xlab = "")
lines(alldata$datetime, alldata$Sub_metering_2, col = "red")
lines(alldata$datetime, alldata$Sub_metering_3, col = "blue")
legend("topright", legend = names(alldata)[6:8], lty = 1,
       col = c("black", "red", "blue"))
# dev.off()
```

```{r plot4}
# make plot 4
# png(paste0(figpath, "plot4.png"))
par(mfrow = c(2, 2))
plot(alldata$datetime, alldata$Global_active_power, type = "l",
     xlab = "", ylab = "Global Active Power")
plot(alldata$datetime, alldata$Voltage, type = "l",
     xlab = "datetime", ylab = "Voltage")
plot(alldata$datetime, alldata$Sub_metering_1, type = "l",
     xlab = "", ylab = "Energy sub metering")
lines(alldata$datetime, alldata$Sub_metering_2, col = "red")
lines(alldata$datetime, alldata$Sub_metering_3, col = "blue")
legend("topright", legend = names(alldata)[6:8], lty = 1,
     col = c("black", "red", "blue"), bty = "n")
plot(alldata$datetime, alldata$Global_reactive_power, type = "l",
     xlab = "datetime", ylab = names(alldata)[3])
# dev.off()
```
